---
title: "some `rstan` trials"
author: "Christof Neumann"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(size = 'footnotesize', fig.align = 'center', echo = TRUE)
options(width = 118)
```

```{r, message=FALSE, warning=FALSE}
library(rstan)
library(bayesplot)
```

# simple linear model

```{r, out.width="50%"}
xdata <- data.frame(x1 = rnorm(200))
xdata$resp <- -3 + xdata$x1 * 1.2 + rnorm(nrow(xdata), sd = 4)
plot(xdata$x1, xdata$resp)
```

For `rstan` we need the data in `list` format:
```{r}
xd <- list(N = nrow(xdata), x1 = xdata$x1, resp = xdata$resp)
```

We can specify the model as a character string, but need to pass this as `model_code=` argument in the `stan()` function. 

Technically, `stan` developers recommend to write these stan models into external files (blabla.stan). In the subsequent chunk that runs the model, we would need to use the `stan()` function (based on the blabla.stan file).

To make things even more complicated, here we use a `knitr` specific tool, the `stan` engine. So we can spell out the model directly in `stan` code, assign it a name and even cache it (so the compiled version is stored unless the chunk is modified). In this first example, the model is called `mod1` although there is no way of knowing that except looking at the source .Rmd file.

```{r stanmod1, engine='stan', output.var="mod1", cache=TRUE}
data {
  int<lower=0> N;
  vector[N] x1;
  vector[N] resp;
}
parameters {
  real intercept;
  real slope1;
  real<lower=0> sigma;
} 
model {
  intercept ~ normal(-3, 1);
  slope1 ~ normal(1.2, 1);
  sigma ~ exponential(1);
  resp ~ normal(intercept + slope1 * x1, sigma);
}
```

Now we can fit the model to our data:

```{r, results='hide'}
fit1 <- sampling(mod1, data = xd, iter = 5000)
```

```{r}
print(fit1)
coef(lm(xd$resp ~ xd$x1))
```

So this worked well. Now let's set one of the priors completely wrong. The intercept we used to generate the data was negative, so let's constrain the prior for this one to positive values.

```{r stanmod1b, engine='stan', output.var="mod1b", cache=TRUE}
data {
  int<lower=0> N;
  vector[N] x1;
  vector[N] resp;
}
parameters {
  real intercept;
  real slope1;
  real<lower=0> sigma;
} 
model {
  intercept ~ exponential(1);
  slope1 ~ normal(1.2, 1);
  sigma ~ exponential(1);
  resp ~ normal(intercept + slope1 * x1, sigma);
}
```

```{r, results='hide', warning = FALSE}
fit1b <- sampling(mod1b, data = xd, iter = 5000)
```

```{r}
print(fit1b)
```

So this behaved as promised, i.e. the intercept point estimate for the intercept is positive (just quite).

# multiple linear regression with interaction

```{r}
xdata <- data.frame(x1 = rnorm(500), x2 = rnorm(500))
xdata$resp <- -3 + 
  0.3 * xdata$x1 + 
  (-0.5) * xdata$x2 +
  1.2 * xdata$x1 * xdata$x2 +
  rnorm(nrow(xdata), sd = 4)
par(mfrow = c(1, 3))
plot(xdata$x1, xdata$resp)
plot(xdata$x2, xdata$resp)
plot(xdata$x1 * xdata$x2, xdata$resp)

xd <- list(N = nrow(xdata), x1 = xdata$x1, x2 = xdata$x2, resp = xdata$resp)
```

Interactions need to be coded 'manually', i.e. the interaction term has to be created in the `transformed data` block. For multiplication, it apparently is necessary to use "`.*`".

```{r stanmod2, engine='stan', output.var="mod2", cache=TRUE}
data {
  int<lower=0> N;
  vector[N] x1;
  vector[N] x2;
  vector[N] resp;
}
transformed data {
  vector[N] inter;
  inter = x1 .* x2;
}
parameters {
  real intercept;
  real slope1;
  real slope2;
  real slope12;
  real<lower=0> sigma;
} 
model {
  intercept ~ normal(-3, 1);
  slope1 ~ normal(1.2, 1);
  slope2 ~ normal(1.2, 1);
  slope12 ~ normal(1.2, 1);
  sigma ~ exponential(1);
  resp ~ normal(intercept + slope1 * x1 + slope2 * x2 + slope12 * inter, sigma);
}
```


```{r, results='hide', warning = FALSE}
fit2 <- sampling(mod2, data = xd, iter = 5000)
```

```{r}
print(fit2)
coef(lm(resp ~ x1 * x2, data = xd))
```

```{r, out.width="50%"}
stan_dens(fit2, "slope12")
```

```{r, out.width="50%"}
mcmc_areas(x = as.matrix(fit2), pars = "slope12", prob = 0.5)
```



