---
title: "MMI and null models"
author: "Christof Neumann"
date: "1/3/2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(MuMIn)
```

This litte exercise looks at the consequences of model selection in the context of data dredging. By dredging I mean here that given a set of predictors all possible subsets of models are fitted (excluding interactions).

The idea is to generate data sets of completely random numbers with varying numbers of predictors. For a given data set all possible models are fitted using the `dredge()` function of the `MuMIn` package.

```{r, message = FALSE}
set.seed(123)
# number of replicates
replicates <- 500
# results backbone
res <- data.frame(replicate = 1:replicates, npred = NA, 
                  weightnull = NA, deltanull = NA, nmod = NA)
# number of predictors (between 1 and 10)
res$npred <- sample(1:10, nrow(res), replace = TRUE)
# number of observations per data set
N <- 200 

for (i in 1:nrow(res)) {
  npred <- res$npred[i]
  # generate response and X predictors
  xdata <- data.frame(matrix(rnorm(n = N * (npred + 1)), nrow = N))
  colnames(xdata) <- c("resp", paste0("var", 1:npred))
  # create model formula for full model
  xform <- as.formula(paste("resp ~", paste0("var", 1:npred, collapse = "+")))
  # fit full model
  fullmod <- lm(formula = xform, data = xdata, na.action = "na.fail")
  # dredge
  x <- data.frame(dredge(fullmod))
  # extract delta and weight for null
  res$deltanull[i] <- x["1", "delta"]
  res$weightnull[i] <- x["1", "weight"]
  # number of fitted models
  res$nmod[i] <- nrow(x)
}
```

```{r}
# summarize data for plots
pdata <- aggregate(res$deltanull, by = list(npred = res$npred), function(X) sum(X > 2)/length(X))
plot(pdata$npred, pdata$x, xlab = "number of predictors", ylab = "proportion of model sets with null outside delta 2", las = 1)
mtext(text = table(res$npred), side = 3, line = 0, at = as.numeric(names(table(res$npred))), cex = 0.5)
```

```{r}
pdata <- aggregate(res$deltanull, by = list(npred = res$npred), mean)
plot(pdata$npred, pdata$x, xlab = "number of predictors", ylab = "mean delta of null", las = 1)
mtext(text = table(res$npred), side = 3, line = 0, at = as.numeric(names(table(res$npred))), cex = 0.5)
```

```{r}
pdata <- aggregate(res$weightnull, by = list(npred = res$npred), mean)
plot(pdata$npred, pdata$x, xlab = "number of predictors", ylab = "mean weight of null", las = 1)
mtext(text = table(res$npred), side = 3, line = 0, at = as.numeric(names(table(res$npred))), cex = 0.5)
```

